### 2020-12-24 paper/sex_selection_spacing-ver5.tex
Additional edits

[HERE!]

The use of sex selection, and therefore the length of birth intervals, increases with
lower desired fertility and with parity, provided the desired number of sons has 
not been reached \citep{Portner2015b,Jayachandran2017}.

% Factors/predictions
I use female education and area of residence as the two principal explanatory variables in
the empirical analyses because those are the two closest proxies available in the data to
cover the explanations above. 
Prior research shows that both higher female education and urbanization are associated with 
lower fertility and increased sex selection use, and, therefore, likely also changes birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Furthermore, both variables are of interest, given the tremendous changes in India. 
Figure \ref{fig:education_over_time} shows the substantial increase in female education,
and the urban proportion of the population almost doubled from 18\% in 1960 to 35\% in 
2019 \citep{United-Nations2019}.

@@

% [Parity progression probability]
The parity progression and the sex ratio show two broad trends.
First, 
in line with the falling total fertility rate, the likelihood of a next birth 
has decreased over time.
The likelihood of a next birth fell more rapidly, the higher the education, the higher the 
parity, and with at least one son.
Within a given spell and period, parity progressions are lower in urban than rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Second, 
the spread of sex selection shows clearly in the sex ratios of next births, which has 
become more male-dominated for women with no sons.
The percentage of births that were boys increased more quickly, the higher the education 
and the higher the parity.
There are no clear trends for the other sex compositions.
Within a given spell and period combination and in the absence of a son, sex ratios are higher 
the more educated the mother is and with higher parity. 
Sex ratios are also higher in urban than in rural areas.
Some women with one son also show an unnaturally high percentage of boys, although the 
failing fertility makes these estimates noisy.

@@

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months from the prior birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that repeated abortions are associated with higher infant mortality
for the child eventually born; boys born after long birth intervals to families with
well-educated mothers, only daughters, and high predicted sex ratios---the combination of
which suggests repeated sex-selective abortions---have no higher infant mortality than other
children.

@@

Second, what is the relationship between female labor force participation and sex selection? 
Women may be staying out of the labor market precisely because sex selection makes 
them more likely to have a boy and increases the expected birth spacing. 
Better job opportunities for women could reduce sex selection use for two reasons. 
First, it becomes more expensive to be out of the labor market for long periods. 
Second, the differential in potential earnings between husband and wife would decline, 
making it more attractive to invest in daughters' human capital.  
This approach could, however, be a double-edged sword. 
If better job opportunities further lower fertility, the use of sex selection may increase, 
everything else being equal.
Understanding the trade-off between long-term benefits from improvements in women's labor 
force participation and short-term costs from potential increases in sex selection is of 
paramount importance.

>>>>

% Factors/predictions
The closest available proxies in the data for capturing the explanations linking
fertility and birth spacing are the sex composition of previous children, mothers'
education, and area of residence. 
The remainder of the section discusses how the three variables tie into the
explanations and the predicted effects.

Sex composition proxies for son preference and the use of
sex selection, both unobserved.%
\footnote{
The surveys ask about the ideal number of sons and 
daughters, but these appear to be unreliable son 
preference indicators; better-educated women simultaneously show 
declining son preference and increasing sex selection 
\citep{bhat03,pande07}.
} 
In the absence of sons, sex selection appears to increase 
with lower desired fertility and higher parity
\citep{Portner2015b,Jayachandran2017}. 
Thus, for a given parity, birth intervals should lengthen 
substantially over time for women with no sons, lower 
desired fertility, and better access to sex selection.

Both higher female education and urbanization are associated with 
lower fertility and increased sex selection use, and, therefore, likely also changes in birth spacing
\citep{das_gupta97,retherford03b,Portner2015b}.
Furthermore, both variables are representative of
the tremendous changes in India. 
Figure \ref{fig:education_over_time} shows the substantial increase in female education,
and the urban proportion of the population almost doubled from 18\% in 1960 to 35\% in 
2019 \citep{United-Nations2019}.

@@

% [Parity progression probability]
The parity progression and the sex ratio show two broad trends.
First, 
in line with the falling total fertility rate, the likelihood of a next birth 
has decreased over time.
The likelihood of a next birth fell more rapidly, the higher the education and the 
higher the parity.
Within a given spell and period, parity progressions are lower in urban than rural areas, 
if at least one son is present, and the more educated the mother.

% [Sex ratios]
Second, sex ratios have become more male-dominated for women with no sons, indicating the spread 
of sex selection. 
The percentage of boys increased more quickly, the higher the education and the higher the parity. 
There are no clear trends for the other sex compositions. 
Within a given spell and period combination and in the absence of a son, sex ratios are higher, 
the more educated the mother, and the higher the parity. Sex ratios are also higher in urban than 
in rural areas.
Some women with one son also show an unnaturally high percentage of boys, although the 
failing fertility makes these estimates noisy.

@@

% Mortality
There has been a substantial reduction in infant mortality over time, and the size of the 
reductions is inversely related to the mother's education.
Hence, there is now little difference in mortality risk across education groups if the 
birth took place more than 21 months after the previous birth.
Short birth spacing is still associated with higher mortality, although the effect is 
small for the best-educated women.
There is no evidence that repeated abortions are associated with higher infant mortality
for the child eventually born; boys born after long birth intervals to families with
well-educated mothers, only daughters, and high predicted sex ratios---the combination of
which suggests repeated sex-selective abortions---have no higher infant mortality than other
children.

@@

Second, what is the relationship between female labor force participation and sex selection? 
Women may be staying out of the labor market precisely because sex selection makes 
them more likely to have a boy and increases the expected birth spacing. 
Better job opportunities for women could reduce sex selection use for two reasons. 
First, it becomes more expensive to be out of the labor market for long periods. 
Second, the differential in potential earnings between husband and wife would decline, 
making it relatively more attractive to invest in daughters' human capital.  
This approach could, however, be a double-edged sword. 
If better job opportunities further lower fertility, the use of sex selection may increase, 
everything else being equal.
Understanding the trade-off between long-term benefits from improvements in women's labor 
force participation and short-term costs from potential increases in sex selection is of 
paramount importance.
